---
layout: post
title: "Gay Marriage and Abortion"
date: 2017-01-21
output:
  html_document:
    css: style.css
share: true
categories: blog
excerpt: "Longitudinal Analysis of Gay Marriage and Abortion"
tags: [rstats]
---

```{r message=FALSE, warning=FALSE, echo=FALSE}
library(tidyverse)
library(haven)
library(car)
library(reshape2)
library(plotly)
library(extrafont)
library(extrafontdb)
library(RColorBrewer)

Sys.setenv("plotly_username"="ryanburge")
Sys.setenv("plotly_api_key"="bvzv16do7x")

gss <- read_dta("D:/reltrad.dta")
```

## Abortion Support Over Time

I don't think I've ever seen anyone do abortion support over time using reltrad. This is all GSS data. The only small hiccup is that they didn't ask abortion questions in 1986. I just smoothed that over in the visuals. 

All the plots are interactive. Mouse over to see that actual value for each year and each tradition. 

```{r message=FALSE, warning=FALSE, echo=FALSE}
df <- select(gss, year, age, reltrad, abdefect, abnomore, abhlth, abpoor, abrape, absingle, abany)
df$abdefect <- as_factor(df$abdefect)
df$abdefect <- as.numeric(df$abdefect)
df$abnomore <- as_factor(df$abnomore)
df$abnomore <- as.numeric(df$abnomore)
df$abhlth <- as_factor(df$abhlth)
df$abhlth <- as.numeric(df$abhlth)
df$absingle <- as_factor(df$absingle)
df$absingle <- as.numeric(df$absingle)
df$abany <- as_factor(df$abany)
df$abany <- as.numeric(df$abany)
df$abrape <- as_factor(df$abrape)
df$abrape <- as.numeric(df$abrape)

df$abdefect[df$abdefect >=3] <- NA
df$abnomore[df$abnomore >=3] <- NA
df$abhlth[df$abhlth >=3] <- NA
df$absingle[df$absingle >=3] <- NA
df$abany[df$abany >=3] <- NA
df$abrape[df$abrape >=3] <- NA

df$abdefect <- recode(df$abdefect, "1=1; 2=0")
df$abnomore <- recode(df$abnomore, "1=1; 2=0")
df$abhlth <- recode(df$abhlth, "1=1; 2=0")
df$abpoor <- recode(df$abpoor, "1=1; 2=0")
df$absingle <- recode(df$absingle, "1=1; 2=0")
df$abany <- recode(df$abany, "1=1; 2=0")
df$abrape <- recode(df$abrape, "1=1; 2=0")


abort <- df %>% group_by(year, reltrad) %>% summarise(defect = mean(abdefect, na.rm =TRUE), nomore = mean(abnomore, na.rm =TRUE), health = mean(abhlth, na.rm =TRUE), poor = mean(abpoor, na.rm =TRUE), rape = mean(abrape, na.rm =TRUE), single = mean(absingle, na.rm =TRUE), any= mean(abany, na.rm =TRUE))
a2 <- melt(abort, id.vars=c("year", "reltrad"))

a2$reltrad[a2$reltrad ==1] <- "Evangelical"
a2$reltrad[a2$reltrad ==2] <- "Mainline"
a2$reltrad[a2$reltrad ==3] <- "Black Protestant"
a2$reltrad[a2$reltrad ==4] <- "Catholic"
a2$reltrad[a2$reltrad ==5] <- "Jewish"
a2$reltrad[a2$reltrad ==6] <- "Other"
a2$reltrad[a2$reltrad ==7] <- "No Religion"
a2$reltrad <- as_factor(a2$reltrad)


Nth.delete<-function(dataframe, n)dataframe[-(seq(n,to=nrow(dataframe),by=n)),]
a2 <- Nth.delete(a2, 8)
a2 <- filter(a2, year != 1986)


p1 <- ggplot(a2 %>% filter(variable == "nomore"), aes(x=year, y=value*100, label = reltrad, color =reltrad)) + 
  geom_line(aes(group = reltrad), size=1.5 ) + xlab("Year") + ylab("Percent in Favor of Abortion") + 
  ggtitle("If she is married and does not want any more children?") + labs(colour = "Religious Tradition") +
  theme(legend.position="bottom")  + 
  theme(text=element_text(size=16, family="KerkisSans"))+ ylim(0,100)+  
  scale_colour_brewer("Religious Tradition", palette="Set2") 

p2 <-ggplot(a2 %>% filter(variable == "defect"), aes(x=year, y=value*100, label = reltrad, color =reltrad)) + 
  geom_line(aes(group = reltrad), size=1.5 ) + xlab("Year") + ylab("Percent in Favor of Abortion") + 
  ggtitle("If there is a strong chance of serious defect in the baby?") + labs(colour = "Religious Tradition") +
  theme(legend.position="bottom")  + 
  theme(text=element_text(size=16, family="KerkisSans"))+ ylim(0,100)+  
  scale_colour_brewer("Religious Tradition", palette="Set2") 

p3 <- ggplot(a2 %>% filter(variable == "health"), aes(x=year, y=value*100, label = reltrad, color =reltrad)) + 
  geom_line(aes(group = reltrad), size=1.5 ) + xlab("Year") + ylab("Percent in Favor of Abortion") + 
  ggtitle("If the woman''s own health is seriously endangered by the pregnancy?") + labs(colour = "Religious Tradition") +
  theme(legend.position="bottom")  + 
  theme(text=element_text(size=16, family="KerkisSans"))+ ylim(0,100)+  
  scale_colour_brewer("Religious Tradition", palette="Set2") 

p4 <- ggplot(a2 %>% filter(variable == "poor"), aes(x=year, y=value*100, label = reltrad, color =reltrad)) + 
  geom_line(aes(group = reltrad), size=1.5 ) + xlab("Year") + ylab("Percent in Favor of Abortion") + 
  ggtitle("If the family has a very low income and cannot afford any more children?") + labs(colour = "Religious Tradition") +
  theme(legend.position="bottom")  + 
  theme(text=element_text(size=16, family="KerkisSans"))+ ylim(0,100)+  
  scale_colour_brewer("Religious Tradition", palette="Set2") 

p5 <- ggplot(a2 %>% filter(variable == "rape"), aes(x=year, y=value*100, label = reltrad, color =reltrad)) + 
  geom_line(aes(group = reltrad), size=1.5 ) + xlab("Year") + ylab("Percent in Favor of Abortion") + 
  ggtitle("If she became pregnant as a result of rape?") + labs(colour = "Religious Tradition") +
  theme(legend.position="bottom")  + 
  theme(text=element_text(size=16, family="KerkisSans")) + ylim(0,100)+  
  scale_colour_brewer("Religious Tradition", palette="Set2") 

p6 <- ggplot(a2 %>% filter(variable == "single"), aes(x=year, y=value*100, label = reltrad, color =reltrad)) + 
  geom_line(aes(group = reltrad), size=1.5 ) + xlab("Year") + ylab("Percent in Favor of Abortion") + 
  ggtitle("If she is not married and does not want to marry the man?") + labs(colour = "Religious Tradition") +
  theme(legend.position="bottom")  + 
  theme(text=element_text(size=16, family="KerkisSans"))+ ylim(0,100)+  
  scale_colour_brewer("Religious Tradition", palette="Set2") 

p7 <- ggplot(a2 %>% filter(variable == "any"), aes(x=year, y=value*100, label = reltrad, color =reltrad)) + 
  geom_line(aes(group = reltrad), size=1.5 ) + xlab("Year") + ylab("Percent in Favor of Abortion") + 
  ggtitle("The woman wants it for any reason?") + labs(colour = "Religious Tradition") +
  theme(legend.position="bottom")  + 
  theme(text=element_text(size=16, family="KerkisSans"))+ ylim(0,100) +  
  scale_colour_brewer("Religious Tradition", palette="Set2") 
```

```{r message=FALSE, warning=FALSE, echo=FALSE, fig.width=12}
ggplotly(p1)
ggplotly(p2)
ggplotly(p3)
ggplotly(p4)
ggplotly(p5)
ggplotly(p6)
ggplotly(p7)
```

## Has Pro-Life Support Waned Among Younger Evangelicals

This data is just from 2010-2014. I wanted to get a more contemporary view of evangelicalism. The data is broken down into three basic generations. As you can see, younger evangelicals are as pro-life as middle aged ones. In some ways younger evangelicals are actually more conservative on abortion than middle agers. 

The Catholic trend is fascinating, though. Old Catholics are as conservative as old evangelicals. Less so in the younger age categories. 

```{r message=FALSE, warning=FALSE, echo=FALSE, fig.width=12}
df <- filter(df, year >=2010) 


#df$age <- as_factor(df$age)
#df$age <- as.numeric(df$age)

df <- na.omit(df)

#cut(df$age, breaks =3)
df$gen <- recode(df$age, "17.9:41 =1; 42:65=2; 65:90=3")

a1 <- df %>% group_by(gen, reltrad) %>% summarise(defect = mean(abdefect, na.rm =TRUE), nomore = mean(abnomore, na.rm =TRUE), health = mean(abhlth, na.rm =TRUE), poor = mean(abpoor, na.rm =TRUE), rape = mean(abrape, na.rm =TRUE), single = mean(absingle, na.rm =TRUE), any= mean(abany, na.rm =TRUE))

a1$reltrad[a1$reltrad ==1] <- "Evangelical"
a1$reltrad[a1$reltrad ==2] <- "Mainline"
a1$reltrad[a1$reltrad ==3] <- "Black Protestant"
a1$reltrad[a1$reltrad ==4] <- "Catholic"
a1$reltrad[a1$reltrad ==5] <- "Jewish"
a1$reltrad[a1$reltrad ==6] <- "Other"
a1$reltrad[a1$reltrad ==7] <- "No Religion"
a1$reltrad <- as_factor(a1$reltrad)

ab1 <- melt(a1, id.vars=c("gen", "reltrad"))

ab1$gen[ab1$gen ==1] <- "18-41 years old"
ab1$gen[ab1$gen ==2] <- "42-65 years old"
ab1$gen[ab1$gen ==3] <- "Over 65 years old"

ab1$variable <- factor(ab1$variable, labels = c("Birth Defect", "No More Kids", "Mother's Health", "Poverty", "Rape", "Unmarried", "Any Reason"))



p8 <- ggplot(ab1 %>% filter(gen == "18-41 years old"), aes(x = value*100, y = variable))  +
  geom_point(shape=21, size =4, aes(fill = factor(reltrad))) +  theme(legend.title=element_blank()) +
  theme(legend.position = "bottom")  + scale_fill_brewer(palette = "Set2") + 
  ylab("Abortion Situation") + xlab("Percent in Favor of Abortion") +
  ggtitle("                                             18-41 years old")+ 
  theme(text=element_text(size=16, family="KerkisSans")) + xlim(0,100)

p9 <- ggplot(ab1 %>% filter(gen == "42-65 years old"), aes(x = value*100, y = variable))  +
  geom_point(shape=21, size =4, aes(fill = factor(reltrad))) +  theme(legend.title=element_blank()) +
  theme(legend.position = "bottom")  + scale_fill_brewer(palette = "Set2") + 
  ylab("Abortion Situation") + xlab("Percent in Favor of Abortion") +
  ggtitle("                                             42-65 years old")+ 
  theme(text=element_text(size=16, family="KerkisSans")) + xlim(0,100)


p10 <- ggplot(ab1 %>% filter(gen == "Over 65 years old"), aes(x = value*100, y = variable))  +
  geom_point(shape=21, size =4, aes(fill = factor(reltrad))) +  theme(legend.title=element_blank()) +
  theme(legend.position = "bottom")  + scale_fill_brewer(palette = "Set2") + 
  ylab("Abortion Situation") + xlab("Percent in Favor of Abortion") +
  ggtitle("                                         Over 65 years old")+ 
  theme(text=element_text(size=16, family="KerkisSans")) + xlim(0,100)

ggplotly(p8)
ggplotly(p9)
ggplotly(p10)

```

## Support for Gay Marriage Over Time

I wanted to compare abortion support and gay marriage support. Unfortunately the GSS didn't ask these questions as frequently as the abortion ones.

Notice that the gay marriage question only goes back to 2006 here. But you can see the upward trend easily. 

The second graph has gaps in years where the question wasn't asked. Again, the upward trend. 

```{r message=FALSE, warning=FALSE, echo=FALSE, fig.width=12}
df <- select(gss, year, reltrad, marhomo, homosex)
df$marhomo <- as_factor(df$marhomo)
df$marhomo <- as.numeric(df$marhomo)
df$homosex <- as_factor(df$homosex)
df$homosex <- as.numeric(df$homosex)


df$marhomo[df$marhomo >=6] <- NA
df$homosex[df$homosex >=6] <- NA

df$marhomo <- recode(df$marhomo, "1=1; 2=1; 3=0; 4=0; 5=0")
df$homosex <- recode(df$homosex, "4=1; 1=0; 2=0; 3=0; 5=0")



gay <- df %>% group_by(year, reltrad) %>% summarise(marriage = mean(marhomo, na.rm =TRUE), gaysex = mean(homosex, na.rm =TRUE))
a2 <- melt(gay, id.vars=c("year", "reltrad"))

a2$reltrad[a2$reltrad ==1] <- "Evangelical"
a2$reltrad[a2$reltrad ==2] <- "Mainline"
a2$reltrad[a2$reltrad ==3] <- "Black Protestant"
a2$reltrad[a2$reltrad ==4] <- "Catholic"
a2$reltrad[a2$reltrad ==5] <- "Jewish"
a2$reltrad[a2$reltrad ==6] <- "Other"
a2$reltrad[a2$reltrad ==7] <- "No Religion"
a2$reltrad <- as_factor(a2$reltrad)

a2 <- Nth.delete(a2, 8)

p11 <- ggplot(a2 %>% filter(variable == "marriage"), aes(x=year, y=value*100, label = reltrad, color =reltrad)) + 
  geom_line(aes(group = reltrad), size=1.5 ) + xlab("Year") + ylab("Percent Who Strongly Agree or Agree") + 
  ggtitle("Homosexual couples should have the right to marry one another.") + labs(colour = "Religious Tradition") +
  theme(legend.position="bottom")  + 
  theme(text=element_text(size=16, family="KerkisSans"))+ ylim(0,100)+  xlim(2006, 2014) +
  scale_colour_brewer("Religious Tradition", palette="Set2") 

p12 <- ggplot(a2 %>% filter(variable == "gaysex"), aes(x=year, y=value*100, label = reltrad, color =reltrad)) + 
  geom_line(aes(group = reltrad), size=1.5 ) + xlab("Year") + ylab("Not Wrong At All (Percent)") + 
  ggtitle("What about sexual relations between two adults of the same sex") + labs(colour = "Religious Tradition") +
  theme(legend.position="bottom")  + 
  theme(text=element_text(size=16, family="KerkisSans"))+ ylim(0,100) + 
  scale_colour_brewer("Religious Tradition", palette="Set2")

ggplotly(p11)
ggplotly(p12)

```